---
title: "Is it brownian?"
author: "MK Ianni-Ravn"
date: "04/03/2022"
output: html_document
---


```{r}
devtools::load_all("~/slendr")
setup_env()
check_env()
library(ape)
library(ggtree)
library(cowplot)
library(tidyverse)
library(sf)
library(MetBrewer)
```


First, let's set up the simulation. In a real Brownian motion in 2d, particles walk in x and y dimensions, taking steps drawn from a normal distribution of mean 0 and sd 1. Martin and I implemented a new dispersal function `brownian` which is designed to recapitulate this. When making offspring, slendr does the following:

1. Find a mate somewhere within the `mate_dist`.
2. Randomly pick one of the parents and disperse the child independently in x and y following a normal distribution.

Note 1: If the mating distance is small enough, the fact that children can be dispersed from either parent should (hopefully) not be super problematic. But we can check this...
Note 2: there is another parameter called `competition_dist`, which determines within how large of a radius individuals compete for resources. For this purpose it's probably best to use a small competition distance, otherwise the individuals break up into small clusters over generations (see Martin's vignette at https://www.slendr.net/articles/vignette-03-interactions.html).


```{r}
# setting some parameters
disp_f <- "brownian"
comp_d <- 0
mate_d <- 1
disp_d <- 1
# making a landscape...
map <-
  world(
    xrange = c(1, 100),
    yrange = c(1, 100),
    landscape = "blank"
  )
# defining the population
pop <- population(
  "POP",
  time = 1,
  N = 500,
  dispersal_fun = disp_f,
  competition_dist = comp_d ,
  mate_dist = mate_d,
  dispersal_dist = disp_d,
  center = c(50, 50),
  radius = 50,
  map = map
)
```

`
Now, let's make and run out model. Here, I just care about the tree so I'll simulate a non-recombining locus of length 1. I added a new option, `retainall`, which if set to `True` will tell slim to remember every individual in the tree- not just the nodes. This will let us track everyone in a lineage and look at dispersal more directly.

```{r}
# how many gens?
gens <- 500
# make the model
model <- compile(
  populations = pop,
  generation_time = 1,
  sim_length = gens,
  resolution = 0.5
)
# who should we sample?
samples <- sampling(model, times = gens, list(pop, 10))
# run model
slim(
  model,
  sequence_length = 1,
  recombination_rate = 0, # one tree locus
  method = "batch",
  random_seed = 42,
  retainall=TRUE, # a new option
  sampling = samples
)
```

Let's load some stuff up! Question here: are we keeping..... everyone..... because we're not simplifying...?

```{r}
# now let's play with all the remembered nodes
tsu <- ts_load(model) 
nds <- tsu %>% ts_nodes()
edg <- tsu %>% ts_edges() %>% rename("edge_id"="id")
dat <- tsu %>% ts_data()
dat %>% filter(!is.na(name))

dat %>% filter(retained==TRUE)
# extract data

dat <- dat %>%
  rowwise() %>%
  mutate(x=unlist(location)[1],
         y=unlist(location)[2]) %>%
  ungroup()
```

Now we can investigate how the individuals are spreading. A first simple plot:

```{r}
# let's see how we're spread
ggplot(dat)+
  geom_point(aes(x=x,y=y,col=time))+
  theme(legend.position="none") # this looks amazing!
```
Seems to check out! Let's see how they spread along the `x` dimension over time.

```{r}
# let's see how we're spread
ggplot(dat) +
  geom_density(aes(x=x,col=as.factor(time)))+
  theme(legend.position="none") # this looks amazing!
```

This looks pretty good- we would expect there to be a gradual increase in the spread and it's what we're seeing. But let's also take a look at the standard deviations of the positions. 


```{r}
# and the standard devs over time
dat %>% group_by(time) %>%
  summarise(sdx=sd(x),sdy=sd(y)) %>%
  pivot_longer(cols=c("sdx","sdy"),names_to="dim",values_to="sd") %>%
  ggplot(aes(x=time,y=sd,col=dim)) +geom_point(size=0.1)+ geom_line(size=0.1)+geom_smooth(method="lm")
```
Here, I'm wondering why it's so spiky. But overall we see an increase, nice! Let's see how many nodes there are, over time. 

```{r}
dat %>% group_by(time) %>%
  summarise(n=n()) %>%
  ggplot(aes(x=time,y=n))+geom_point()
```
So everyone in the last generation is sampled, but back in time there's an exponential decrease. 

I want to join all the information above to get a working network, where children are connected to their parents. I cannibalised some of Martin's code and tidyverse-d the rest. 

```{r}
# first we get all the nodes which are parents
parent_nodes <- dat %>%
  dplyr::as_tibble() %>%
  dplyr::filter(node_id %in% edg$parent) %>% 
  dplyr::select(parent_pop = pop,
                parent_ind_id = ind_id, 
                parent_node_id = node_id,
                parent_time = time, parent_location = location) %>%
  dplyr::left_join(edg, by = c("parent_node_id" = "parent")) %>%
  dplyr::arrange(parent_node_id)
# next we get all the nodes which are children
branch_nodes <- dat %>%
  dplyr::as_tibble() %>%
  dplyr::filter(node_id %in% edg$child) %>%
  dplyr::select(child_pop  = pop,
                child_ind_id = ind_id,
                child_node_id = node_id,
                child_time = time, child_location = location) %>%
  dplyr::inner_join(parent_nodes, by = c("child_node_id" = "child")) %>% # and smash the two dataframes together
  dplyr::arrange(child_node_id)
# get all the lines connecting the parent and child locations
connections <- branch_nodes %>%
  rowwise() %>%
  mutate(pts=st_union(child_location,parent_location)) %>%
  mutate(connection=st_cast(pts,"LINESTRING")) %>% ungroup()
# let the branches hold the line information
branches <- connections %>%
  sf::st_set_geometry("connection")
# let the connections hold the point position of the individuals
connections <- connections %>% rowwise() %>%
  mutate(edge_gens=child_time-parent_time,
         # calculate the x and y displacements
         disp=st_length(connection),
         x_disp=unlist(child_location)[1]-unlist(parent_location)[1],
         y_disp=unlist(child_location)[2]-unlist(parent_location)[2],
         # and scale everything by generations
         disp_pergen=disp/edge_gens,
         x_disp_pergen=x_disp/edge_gens,
         y_disp_pergen=y_disp/edge_gens) %>%
  ungroup() %>%
  sf::st_set_geometry("child_location")
```

Let's explore this data.

```{r}
branches %>%
  ggplot(aes(col=as.factor(child_time)))+
  geom_sf(alpha=0.5)+
  geom_sf(data=connections,size=0.5)+
  theme(legend.position = NULL) +
  scale_color_manual(values=met.brewer("Hokusai2", n = gens+1))
```
Kind of checks out, the most recent individuals are more spread. We'll come back to a similar visualisation later. 

```{r}
# this however seems strange. Why do the terminal children have more than one connection...?
branches %>%
  ggplot(aes(col=child_time==gens+1))+
  geom_sf(alpha=0.5)+
  geom_sf(data=connections,size=0.5)+
  theme(legend.position = NULL) # as does this!
# here, we would expect edge lengths to be 1, given that we record everyone....
connections %>% ggplot(aes(x=child_time,y=edge_gens))+geom_point() # this looks really bad
# here, we would expect an x=y+1 sort of thing....
connections %>% ggplot(aes(x=child_time,y=parent_time))+geom_point() # this looks really bad and really weird
# here, we would expect no spatial clustering of edge generations- which should be 1 anyway
connections %>% ggplot()+geom_sf(aes(col=edge_gens)) # this looks really bad
# here, we want each child to appear twice, because of diploid-ness
```

Now I want to see one lineage walking in space. 

```{r}
### plotting the paths of a lineage
# extract ancestors
anc <- ts_ancestors(tsu,1) %>%
  select(child_phylo_id=child_id,parent_time,child_time)
# sf black magic
anc <- anc %>%
  rowwise() %>%
  # some displacement calculations again
  mutate(xd=unlist(connection)[2]-unlist(connection)[1],
         yd=unlist(connection)[4]-unlist(connection)[3]) %>%
  ungroup()
# let's see it walk in space...
ggplot(anc)+geom_sf(aes(col=child_time)) # this looks good
```
Do the steps look like we'd expect with a 2d Brownian motion?

```{r}
# and let's see how the dispersals are going!
# predictions- should be normal
brown <- data.frame(norm=rnorm(1000,0,1))
# prepare for plotting...
anc_longer <- anc %>% pivot_longer(cols=c("xd","yd"),
                     names_to = "coord",
                     values_to = "disp")
# plot!
ggplot(anc_longer)+geom_density(aes(x=disp,col=coord))+
  geom_density(data=brown,aes(x=norm,col="2d Normal"),lty=2) # this looks great!
```
Sidenote: if you increase the competition distance, xd and yd start to spread out much more compared to expectation.

Now, I want to see whether this agrees with the tree we get if we simplify the data. 

```{r}
### comparing to the simplified tree
# load the tree sequence and simplify it
ts <- ts_load(model) %>%
  ts_recapitate(Ne = 500,recombination_rate = 0) %>%
  ts_simplify()

tree <- ts %>% ts_phylo(i=1)

# let's check how these agree
ts_ancestors(ts, 1) %>%
  ggplot(aes(col=parent_time)) +
  geom_sf(size=2) +geom_sf(data=anc,lty=2) # woohoo, this works!The two agree
```


Show the convex hull again

```{r}
dat <- dat %>%
  group_by(time) %>%
  summarize(geometry = st_union(location)) %>%
  st_convex_hull() %>% 
  mutate(area = st_area(geometry))
```



```{r}
ggplot(dat,aes(x=time,y=sqrt(area))) +
  geom_smooth(aes(x=time,y=sqrt(area)),method="loess",se=FALSE,alpha=0.1)+
  geom_point(size=0.5) +
  geom_line()+
  theme_classic() +
  ggtitle("Bounding Area- time")
```

```{r}
plot <- ggplot(data=dat) +
  ggtitle("Node Distribution") +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())+
  theme_minimal() +
  theme(legend.position="none")

plot <- plot + geom_sf(alpha=0.1, aes(col=as.factor(time))) +
  scale_color_manual(values=met.brewer("Hokusai2", n = gens+1))

plot
```




